---
title: "Testing Scientific Expectations under Relational Event Models with BFpack"
author: "Joris Mulder"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Testing Scientific Expectations under Relational Event Models with BFpack}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  eval = FALSE,          # Code is shown but not run when knitting.
  message = FALSE,       # Set eval = TRUE to run everything live
  warning = FALSE,       # (fitting the REMs over a moving window takes a few minutes).
  collapse = TRUE,
  comment = "#>"
)
```

> **How to use this document.** Every code block below is ready to paste into an
> R session and run in order. To keep the handout light, the chunks are *not*
> executed while knitting (`eval = FALSE` in the setup chunk); instead the
> expected results are shown and discussed in the text. To run the analysis
> live, either flip `eval = FALSE` to `eval = TRUE`, or simply copy each block
> into R. The single full-sequence fit is fast; the moving-window loop refits
> the REM many times and takes a few minutes.

*This is the relational-event companion to the ERGM tutorial. It reuses the same
four-step `BFpack` workflow — fit, get names, formulate constraints, test — on a
relational event model (REM) of militarized interstate disputes. The only real
difference is the last step: for a REM we call `BF()` on the **estimates and
their covariance matrix** (`BF.default`) rather than on a fitted model object.*

---

# Introduction and learning objectives

A **relational event model (REM)** describes a stream of time-stamped, directed
events — *who* acts on *whom*, and *when* — through the *rate* at which each
possible dyad produces the next event. This tutorial teaches **Bayesian
hypothesis testing with Bayes factors and posterior probabilities** for REM
coefficients, using the `BFpack` package together with the `remverse`
(`remify`, `remstats`, `remstimate`).

Why bother going beyond the *p*-value? The same three arguments from the ERGM
tutorial apply — a *p*-value cannot support a null, it is inconsistent under the
null, and it cannot directly test order constraints — but REMs add a bonus:

> Because the sample size of a REM is the **number of events** (not the number
> of nodes), evidence genuinely accumulates as we observe the sequence for
> longer. A Bayes factor is *consistent*, so "collect more events and update the
> Bayes factor" is a legitimate and practical strategy — and, as we will see, we
> can even track how the evidence **changes over time**.

By the end you will be able to:

- fit a tie-oriented REM with `remstats` + `remstimate` and read off the exact
  coefficient names with `coef()`;
- translate a substantive expectation into a `BFpack` hypothesis string using
  equality (`=`) and order (`>`, `<`) constraints;
- compute Bayes factors and posterior probabilities with `BF()` via the
  `BF.default` route (estimates + covariance matrix + `n`);
- run a **moving-window** analysis to see how the evidence for competing
  hypotheses evolves across the observation period; and
- adapt the workflow to your own constrained hypotheses (the exercise).

---

# Setup: packages and data

```{r packages}
# Install once, if needed:
# install.packages("remverse")

library("remverse")
library("BFpack")      # Bayes factors for constrained hypotheses
```

The empirical case is the sequence of **militarized interstate disputes (MIDs)**
from the Correlates of War project (Singer, 1972): a directed event
$i \rightarrow j$ means state $i$ initiated a militarized action against state
$j$. The workshop data folder ships one edgelist plus a set of (time-varying)
country covariates as CSV files.

```{r data}
# Point this at the folder where you unzipped the workshop data.
data_dir <- "dir"
cov <- function(f) read.csv(file.path(data_dir, f), stringsAsFactors = FALSE)

# Edgelist: one row per event, ordered in time.
el <- cov("edgelist_full.csv")
el$time <- as.Date(el$time)
el <- el[order(el$time), ]
```

The sequence contains roughly **4,670 directed events** at **2,429 unique time
points** (several disputes can fall on the same date), spanning the post-war
period. A relational event analysis of these data was presented by Lakdawala et
al. (2026).

## Build the relational event history

`remify()` turns the raw edgelist into a relational event history object. We use
a **tie-oriented** model with an *active-saturated* risk set (every ordered pair
of observed actors is at risk at every time point).

```{r remify}
reh_tie <- remify(el[, c("time", "actor1", "actor2")],
                  directed = TRUE, model = "tie",
                  time_units = "years", riskset = "active_saturated")
reh_tie          # a short summary; reh_tie$M is the number of unique time points
# plot(reh_tie)  # optional: event-rate diagnostics
```

## Actor covariates, including the time-varying regime dummies

The covariates are read as data frames on `remify`'s time scale. The two we will
test are **regime-type dummies**, built from each state's Polity score `dem`
*in each year*, so they **vary over time**:

```{r covariates}
origin   <- min(el$time)                                   # first event = time 0
to_years <- function(d) as.numeric(as.Date(d) - origin) / 365.25

ac <- list(major    = cov("cov_major.csv"),
           dem      = cov("cov_dem.csv"),
           religion = cov("cov_religion.csv"))

maj <- ac$major                                              # 'major'  (0/1)
maj$time  <- to_years(maj$date)

reg <- ac$dem
reg$time  <- to_years(reg$date)
reg$democracy <- as.integer(reg$dem >=  6)                   # democracy dummy
reg$autocracy <- as.integer(reg$dem <= -6)                   # autocracy dummy

rel <- ac$religion                                           # 'islam' / 'christianity'
rel$time  <- to_years(rel$date)
```

> **Read the dummies carefully.** `democracy = 1` when Polity $\ge 6$,
> `autocracy = 1` when Polity $\le -6$; everything in between — the *anocracies* /
> mixed regimes — is the **baseline**. So `send_democracy` and `send_autocracy`
> are each measured *relative to that baseline*, and because Polity changes over
> the decades a state can move between categories across the sequence.

---

# A tie-oriented REM: the four-step workflow

## Step 1 — Fit the full model

The `send()` effects capture how an attribute of the **initiator** shifts its
propensity to start a dispute; `inertia` and `reciprocity` are endogenous
controls (standardized).

```{r fit}
tie_effects <- ~
  inertia(scaling = "std") + reciprocity(scaling = "std") +  # controls
  send("major",        attr_actors = maj) +   # -> send_major
  send("autocracy",    attr_actors = reg) +   # -> send_autocracy
  send("democracy",    attr_actors = reg) +   # -> send_democracy
  send("islam",        attr_actors = rel) +   # -> send_islam
  send("christianity", attr_actors = rel)     # -> send_christianity

training   <- 500                             # burn-in for the endogenous stats
stats_tie  <- remstats(reh_tie, tie_effects = tie_effects, first = training)
fit_tie    <- remstimate(reh_tie, stats_tie)
summary(fit_tie)
```

## Step 2 — Read the exact parameter names

**This step is essential.** The names `BFpack` expects in a hypothesis string
are exactly the names printed by `coef()`.

```{r estimates}
est <- coef(fit_tie)
print(est)
#> Effect names include, among others:
#>   baseline, inertia, reciprocity,
#>   send_major, send_autocracy, send_democracy, send_islam, send_christianity
```

## Step 3 — Formulate competing constrained hypotheses

The two effects of interest are `send_democracy` and `send_autocracy`. Each
competing hypothesis encodes a substantive theory from international relations
about how regime type drives who *initiates* conflict:

- **H1 (null):** `send_democracy = send_autocracy = 0` — *regime type is
  irrelevant*; initiation is driven by other factors.
- **H2 (democratic peace vs. autocratic aggression):**
  `send_democracy < 0 < send_autocracy` — democracies initiate less, autocracies
  more.
- **H3 (autocratic aggression only):** `send_democracy = 0 < send_autocracy` —
  democracies behave like the baseline; autocracies stand out.
- **H4 (democratic restraint only):** `send_democracy < 0 = send_autocracy` —
  democracies are distinctively peaceful; autocracies look like the baseline.

`BF()` adds the **complement** (none of the above) automatically, so we compare
five hypotheses in total.

```{r hypotheses}
hyp_gov <- "send_democracy = send_autocracy = 0;
            send_democracy < 0 < send_autocracy;
            send_democracy = 0 < send_autocracy;
            send_democracy < 0 = send_autocracy"
```

## Step 4 — Compute Bayes factors and posterior probabilities

For a REM there is no dedicated `BF()` method, so we use the general
**`BF.default`** route: supply the estimates `x`, their covariance matrix
`Sigma`, and the number of observations `n`. `BFpack` then builds an *adjusted
fractional Bayes factor* using a Gaussian approximation to the likelihood (Gu et
al., 2017) — the same fast, approximate Bayes factor used for the ERGM's speed
route.

```{r bf}
BF1 <- BF(x = est, Sigma = fit_tie$vcov, n = fit_tie$df.null,
          hypothesis = hyp_gov)
BF1
BF1$PHP_confirmatory
summary(BF1)   # adds the exploratory Pr(=0 / <0 / >0) table for every parameter
```

## Reading the result

The confirmatory posterior probabilities are:

```
Posterior probabilities of the hypotheses
   Pr(hypothesis|data)
H1               0.000    # regime type irrelevant (null)
H2               0.620    # democracy < 0 < autocracy
H3               0.379    # democracy = 0 < autocracy
H4               0.000    # democracy < 0 = autocracy
H5               0.001    # complement
```

Almost all posterior mass sits on **H2 (.62)** and **H3 (.38)**. Both agree that
**autocracies initiate more** than the baseline; the data are *split* on whether
democracies are *additionally* restrained (H2) or simply *indistinguishable from
the baseline* (H3). The Bayes factor between them is only about **1.6** in favor
of H2 — positive but far from decisive. The null H1 and the "restraint-only"
theory H4 are effectively ruled out.

The exploratory part of `summary(BF1)` tells the same story at the parameter
level: `send_autocracy` is confidently positive, whereas `send_democracy` splits
its mass between "= 0" and "< 0" — which is exactly why H2 and H3 both survive.

### Changing the prior probabilities of hypotheses

Equal prior probabilities are the default. To encode differing a-priori
plausibility, pass one weight per hypothesis (including the complement) via
`prior.hyp.conf`:

```{r priors}
BF1b <- BF(x = est, Sigma = fit_tie$vcov, n = fit_tie$df.null,
           hypothesis = hyp_gov,
           prior.hyp.conf = c(1, 1, 1, 1, 2))  # more prior mass on the complement
BF1b$PHP_confirmatory
```

The Bayes factors (the evidence *in the data*) do not change; only the posterior
probabilities, which combine evidence with the prior odds, shift.

---

# Does the evidence hold over time? A moving-window analysis

The test above averages over the whole post-war period. Because a REM's sample
size is the number of *events*, we can instead slide a **window** across the
sequence, refit the REM in each window, and watch the posterior probabilities
evolve (Mulder & Leenders). This turns a single verdict into a *time series of
evidence*.

```{r moving-window}
length.window <- 500                 # events per window
M     <- reh_tie$M                    # number of unique time points
steps <- seq(0, M - training - length.window, by = 100)

php_mat    <- matrix(NA_real_, nrow = length(steps), ncol = 5)
est_dem_w  <- rep(NA_real_, length(steps))
est_auto_w <- rep(NA_real_, length(steps))
mid_time   <- as.Date(rep(NA, length(steps)))
utimes     <- sort(unique(el$time))  # calendar date of each time point

for (i in seq_along(steps)) {
  step    <- steps[i]
  stats_w <- remstats(reh_tie, tie_effects = tie_effects,
                      first = training + step,
                      last  = training + step + length.window)
  fit_w   <- remstimate(reh_tie, stats_w)

  est_w             <- coef(fit_w)                 # use THIS window's estimates
  est_dem_w[i]      <- est_w["send_democracy"]
  est_auto_w[i]     <- est_w["send_autocracy"]

  BF_w <- tryCatch(
    BF(x = est_w, Sigma = fit_w$vcov, n = fit_w$df.null, hypothesis = hyp_gov),
    error = function(e) NULL)
  if (!is.null(BF_w)) php_mat[i, ] <- BF_w$PHP_confirmatory

  mid_idx     <- round(training + step + length.window / 2)
  mid_time[i] <- utimes[mid_idx]                   # window mid-point as a date
}

colnames(php_mat) <- c("H1: none = 0", "H2: dem<0<auto",
                       "H3: dem=0<auto", "H4: dem<0=auto", "complement")
```

Plot the posterior probabilities against the window mid-point date, and — so the
shifts can be read against the effect sizes that drive them — the underlying
estimates on the same time axis:

```{r moving-window-plot}
par(mar = c(1, 1, 1, 1))
par(mfrow = c(2, 1))

matplot(mid_time, php_mat, type = "l", lty = 1, lwd = 2, col = seq_len(5),
        ylim = c(0, 1), xaxt = "n",
        xlab = "Window mid-point (year)", ylab = "Posterior probability",
        main = sprintf("PHPs over a moving window (%d events)", length.window))
axis.Date(1, at = pretty(mid_time), format = "%Y")
legend("topleft", legend = colnames(php_mat), col = seq_len(5),
       lty = 1, lwd = 2, cex = 0.8, bty = "n")

matplot(mid_time, cbind(est_dem_w, est_auto_w), type = "l", lty = 1, lwd = 2,
        col = c("blue", "red"), xaxt = "n",
        xlab = "Window mid-point (year)", ylab = "Estimate",
        main = "send_democracy / send_autocracy over the moving window")
axis.Date(1, at = pretty(mid_time), format = "%Y")
abline(h = 0, lty = 2, col = "grey50")
legend("topleft", legend = c("send_democracy", "send_autocracy"),
       col = c("blue", "red"), lty = 1, lwd = 2, cex = 0.8)
```

The support for H2 versus H3 is **not constant**: in periods where the
`send_democracy` estimate dips clearly below zero, H2 (democratic restraint
*and* autocratic aggression) gains; when it hovers near zero, H3 (autocratic
aggression only) takes over. Aggregating over the whole period hides this
structure — which is the point of quantifying evidence in windows.

---

# Exercise: test your own constrained hypotheses

The regime-type effects invite other substantive questions. Try replacing
`hyp_gov` with each of the following and re-running Step 4 (and, if you like, the
moving window). Predict the outcome first, then check.

1. **Is autocratic aggression *stronger*, *equal*, or *lower* than democratic effect?** Test the *relative magnitudes* directly:

    ```{r ex1}
    hyp_mag <- "send_democracy < send_autocracy;
                send_democracy = send_autocracy;
                send_democracy > send_autocracy"
    ```

Perform the test. Which prior probabilities for the hypotheses did you use?

2. **Different mechanisms.** Can we test the two endogenous effects `inertia` and
`reciprocity` directly against each other (in other words are they on the same scale?).
Formulate different hypotheses on these effects and test them.

3. **An "approximate-equality" (interval) hypothesis.** Ask whether democracy and
   autocracy effects are *practically equivalent* rather than exactly equal,
   using an interval constraint. For instance, we could test that the difference
   does not exceed 0.2. Although the use of `abs()` is not allowed in the
   `hypothesis` argument, it is possible to write, say, `abs(send_democracy - send_autocracy) < 0.2`,
   as two one-sided constraints which both need to be satisfied: 
   `send_democracy - send_autocracy < 0.2 & send_autocracy - send_democracy < 0.2`.
   Try this. Compare the result with the test of the exact equality of the two effects.

For each, report the Bayes factors *and* the posterior probabilities, and state
the prior probabilities you used.

---

# Reporting and interpretation checklist

When you write up a `BFpack` REM analysis, report:

- the **REM specification** (endogenous statistics, covariates, risk set,
  time scale) and how any covariates were constructed — in particular that the
  regime dummies are **time-varying**;
- the exact **hypotheses** in constraint notation, with their substantive
  meaning;
- the **Bayes factors** between hypotheses and the **posterior probabilities**,
  stating the prior probabilities used;
- for a moving-window analysis, the **window length, step, and burn-in**, and
  the time axis used; and
- that `BF.default` uses an **adjusted fractional Bayes factor with a Gaussian
  approximation**, so reported values are approximate.

Remember that posterior probabilities live on a **continuous** scale — no
accept/reject decision is required — and if you *do* commit to a hypothesis,
$1 - P(H \mid \text{data})$ is the conditional probability of being wrong given
these events and this set of hypotheses.

---

# References

Gu, X., Mulder, J., & Hoijtink, H. (2018). Approximated adjusted fractional
Bayes factors: A general method for testing informative hypotheses.
*British Journal of Mathematical and Statistical Psychology, 71*(2), 229–261.

Lakdawala, R., Leenders, R., & Mulder, J. (2026). Not all bonds are created
equal: Dyadic latent class models for relational event data. *Social Networks, 87*,
135–151.

Meijerink-Bosman, M., Back, M., Geukes, K., Leenders, R., & Mulder, J. (2023).
Discovering trends of social interaction behavior over time: An introduction to
relational event modeling. *Behavior Research Methods, 55*, 997–1023.

Mulder, J., & Leenders, R. T. A. J. (2019). Modeling the evolution of
interaction behavior in social networks: A dynamic relational event approach for
real-time analysis. *Chaos, Solitons & Fractals, 119*, 73–85.

Mulder, J., et al. (2021). BFpack: Flexible Bayes factor testing of scientific
expectations in R. *Journal of Statistical Software,* 1–63.

Singer, J. D. (1972). The "Correlates of War" project: Interim report and
rationale. *World Politics, 24*(2), 243–270.

---

```{r session-info, eval=TRUE, echo=FALSE}
# Uncomment when running live to record the software environment:
# sessionInfo()
```
